---
title: "Get model predictions and plot with `ggplot2`"
author: "Stefano Coretta"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get model predictions and plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "300px", fig.align = "center", dpi = 300
)
library(tidyverse)
theme_set(theme_bw())
library(mgcv)
library(tidymv)
```

While `plot_smooths()` offers a streamlined way of plotting predicted smooths from a GAM model, it is too constrained in other more complex cases.

The most general solution is to get the predicted values of the outcome variable according to all the combinations of terms in the model and use this dataframe for plotting.
This method grants the user maximum control over what can be plotted and how to transform the data (if necessary).

I will illustrate how to use the function `predict_gam()` to create a prediction dataframe and how this dataframe can be used for plotting different cases.

## Smooths

First of all let's generate some simulated data and create a GAM model with a factor `by` variable.

```{r model}
library(mgcv)
set.seed(10)
data <- gamSim(4, 400)

model <- gam(
  y ~
    fac +
    s(x2, by = fac),
  data = data
)

summary(model)
```

We can extract the predicted values with `predict_gam()`.

```{r model-p}
model_p <- predict_gam(model)
model_p
```

Now plotting can be easily done with `ggplot2`.
The function `geom_smooth_ci()` can be used to plot the predicted smooths with confidence intervals.

```{r model-plot}
model_p %>%
  ggplot(aes(x2, fit)) +
  geom_smooth_ci(fac)
```

## Surface smooths

Now let's try plotting a model that has a tensor product interaction term (`ti()`).

```{r model-2}
model_2 <- gam(
  y ~
    s(x2) +
    s(f1) +
    ti(x2, f1),
  data = data
)

summary(model_2)
```

Let's get the prediction dataframe and produce a contour plot.

```{r model-2-p}
model_2_p <- predict_gam(model_2)
model_2_p
```

```{r model-2-plot}
model_2_p %>%
  ggplot(aes(x2, f1, z = fit)) +
  geom_raster(aes(fill = fit)) +
  geom_contour(colour = "white")
```

## Smooths at specified values of a continuous predictor

To plot the smooths across a few values of a continuous predictor, we can use the `values` argument in `predict_gam()`.

```{r model-2-values}
predict_gam(model_2, values = list(f1 = c(0.5, 1, 1.5))) %>%
  ggplot(aes(x2, fit)) +
  geom_smooth_ci(f1)
```

## Exclude terms (like random effects)

It is possible to exclude terms when predicting values by means of the `exclude_terms` argument.
This can be useful when there are random effects, like in the following model.

```{r model-3}
data_re <- data %>%
  mutate(rand = rep(letters[1:4], each = 100), rand = as.factor(rand))

model_3 <- gam(
  y ~
    s(x2) +
    s(x2, rand, bs = "fs", m = 1),
  data = data_re
)

summary(model_3)
```

`exclude_terms` takes a character vector of term names, as they appear in the output of `summary()` (rather than as they are specified in the model formula).
For example, to remove the term `s(x2, fac, bs = "fs", m = 1)`, `"s(x2,fac)"` should be used since this is how the summary output reports this term.

```{r model-3-plot}
predict_gam(model_3, exclude_terms = "s(x2,rand)") %>%
  ggplot(aes(x2, fit)) +
  geom_smooth_ci()
```

Of course, it is possible to plot the predicted values of the random effects.

```{r model-3-rand}
predict_gam(model_3) %>%
  ggplot(aes(x2, fit)) +
  geom_smooth_ci() +
  facet_wrap(~rand)
```
